Comparative RNA-Seq: Signal to Noise Documentation

### 2022-03-14 manuscript.Rmd
Spearman correlation and distance filtering added.
The section to produce the spearman correlation and heatmap were added.
The removal of alignments with a maximum distance > 0.15 and the data
used to determine the appropriate threshold are added.

Extra files (old versions of the rendered markdown) have been removed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggtree)
library(treeio)
library(ape)
library(UpSetR)
library(svglite)
library(ROSE)
library(ggpubr)
library(rstatix)
options(dplyr.summarise.inform = FALSE)
source('~/bin/R/functions.R')

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggtree)
library(treeio)
library(ape)
library(UpSetR)
library(svglite)
library(ROSE)
library(ggpubr)
library(rstatix)
library(gplots)
options(dplyr.summarise.inform = FALSE)
source('~/bin/R/functions.R')

@@

Data sets were filtered by distance < 0.15.

-   Threshold was selected based on the distance between genomes within a family
      -   Filtered at this stage to reduce computational time

```{r, phylogenetic distance, eval=T}
load("~/bin/manuscript/data/allDists.Rda")
maxsGenus <- allDists %>% group_by(Genus.x, Genus.y) %>% 
  summarise(max_dist = max(distance)) %>% 
  mutate(same.taxa = ifelse(Genus.x == Genus.y, T, F))
maxsFamily <- allDists %>% group_by(Family.x, Family.y) %>% 
  summarise(max_dist = max(distance)) %>% 
  mutate(same.taxa = ifelse(Family.x == Family.y, T, F))
maxsOrder <- allDists %>% group_by(Order.x, Order.y) %>% 
  summarise(max_dist = max(distance)) %>% 
  mutate(same.taxa = ifelse(Order.x == Order.y, T, F))
maxsClass <- allDists %>% group_by(Class.x, Class.y) %>% 
  summarise(max_dist = max(distance)) %>% 
  mutate(same.taxa = ifelse(Class.x == Class.y, T, F))
maxsPhylum <- allDists %>% group_by(Phylum.x, Phylum.y) %>% 
  summarise(max_dist = max(distance)) %>% 
  mutate(same.taxa = ifelse(Phylum.x == Phylum.y, T, F))
maxsKingdom <- allDists %>% group_by(Kingdom.x, Kingdom.y) %>% 
  summarise(max_dist = max(distance)) %>% 
  mutate(same.taxa = ifelse(Kingdom.x == Kingdom.y, T, F))
genus1 <- maxsGenus %>% ungroup() %>% filter(same.taxa, max_dist > 0) %>% select(max_dist) %>% mutate(group = "Genus")
family1 <- maxsFamily %>% ungroup() %>% filter(same.taxa, max_dist > 0) %>% select(max_dist) %>% mutate(group = "Family")
order1 <- maxsOrder %>% ungroup() %>% filter(same.taxa, max_dist > 0) %>% select(max_dist) %>% mutate(group = "Order")
class1 <- maxsClass %>% ungroup() %>% filter(same.taxa, max_dist > 0) %>% select(max_dist) %>% mutate(group = "Class")
phylum1 <- maxsPhylum %>% ungroup() %>% filter(same.taxa, max_dist > 0) %>% select(max_dist) %>% mutate(group = "Phylum")
kingdom1 <- maxsKingdom %>% ungroup() %>% filter(same.taxa, max_dist > 0) %>% select(max_dist) %>% mutate(group = "Kingdom")

maxs1 <- genus1 %>% bind_rows(family1, order1, class1, phylum1, kingdom1)
maxs1$group <- factor(maxs1$group, levels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus"))

p <- ggplot() +
geom_boxplot(data = maxs1 , aes(x = group, y = max_dist), fill = NA) + coord_flip() +
  geom_hline(yintercept = 0.15)
p
```

### Correlation heatmap

A spearman correlation matrix was produced comparing multiple metrics for each 
feature. The best performing metrics (measured by comparing correlation with
known sRNAs were kept).

```{r correlation_heat_map, eval=T}
load("~/bin/PhD/Chapter_4/chapter_4_files/randomForestDat.Rda")

dat <- dat %>% select(-ID, -power, -motif_count, -read.counts) %>% unique() %>% 
  mutate(cov.min.eval = -log(cov.min.eval), 
         z_mean = -z_mean,
         z_max = -z_max,
         alifold.score  =-alifold.score,
         mfe.score = -mfe.score,
         alifold_cov_score = -alifold_cov_score)

set.seed(101)
matNames <- c("MFE score", "G+C Percentage", "Evolutionary Distance", 
              "Read depth (max)", "Read depth (mean)",
                        "R-scape covariance (mean score)", 
              "R-scape covariance (max score)", 
              "R-scape covariance (number of significant pairs)", 
              "Motif score (mean)", "Motif score (max)", 
              "Alifold Z score (mean)", "Alifold Z score (max)", 
              "Alifold covariance score", "Alifold score", 
              "Number of sequences in alignment", "Sum of motif scores", 
              "Random", "Known sRNA or RINC")

matrixList <- calculateCorrelationMatrix(dat, matNames)
rhoMatrix <- matrixList$rhoMatrix
sigMatrix <- matrixList$sigMatrix

heatmap.2(rhoMatrix, cellnote=round(rhoMatrix, digits = 2),
          notecex=1.5,notecol="black", col=rev(redblue(40)), 
          density.info="none", trace="none", dendrogram=c("column"), 
          symm=F,symkey=T,symbreaks=T, scale="none", key.title = "", 
          srtRow=45, adjRow=c(0, 1), srtCol=45, adjCol=c(1,1), 
          breaks=(-20:20)/20, margins = c(8, 8), cexRow=1.5, 
          cexCol=1.5,font=1)

if(FALSE){
svglite::svglite(paste0(figure_path, "/heatmap.svg"))
heatmap.2(rhoMatrix, cellnote=sigMatrix,notecex=1.5,notecol="black",
          col=rev(redblue(40)), density.info="none", trace="none", 
          dendrogram=c("column"), symm=F,symkey=T,symbreaks=T, scale="none", 
          key.title = "", srtRow=45, adjRow=c(0, 1), srtCol=45, adjCol=c(1,1), 
          breaks=(-20:20)/20, margins = c(8, 8), cexRow=1.5, cexCol=1.5,font=2)

dev.off()
}
```

```{r subset_correlation_heat_map, eval=T}
load("~/bin/PhD/Chapter_4/chapter_4_files/randomForestDat.Rda")

dat <- dat %>% select(group, 
                      read.max.score, 
                      distance, 
                      alifold_cov_score,
                      cov.min.eval,
                      motif.sum,
                      z_max,
                      mfe.score,
                      gc.score) %>% 
  unique() %>% 
  mutate(cov.min.eval = -log(cov.min.eval), 
         z_max = -z_max,
         mfe.score = -mfe.score,
         alifold_cov_score = -alifold_cov_score)

set.seed(101)
matNames <- c("Read depth (max)", "Evolutionary distance", 
              "Alifold covariance score", "R-scape covariance score", 
              "Sum of motif scores", "Alifold Z score (max)", "MFE score", 
              "G+C Percentage", "Random", "Known sRNA or RINC")

matrixList <- calculateCorrelationMatrix(dat, matNames)
rhoMatrix <- matrixList$rhoMatrix
sigMatrix <- matrixList$sigMatrix

heatmap.2(rhoMatrix, cellnote=round(rhoMatrix, digits = 2),
          notecex=1.5,notecol="black", col=rev(redblue(40)), 
          density.info="none", trace="none", dendrogram=c("column"), 
          symm=F,symkey=T,symbreaks=T, scale="none", key.title = "", 
          srtRow=45, adjRow=c(0, 1), srtCol=45, adjCol=c(1,1), 
          breaks=(-20:20)/20, margins = c(8, 8), cexRow=1.5, 
          cexCol=1.5,font=1)

if(FALSE){
svglite::svglite(paste0(figure_path, "/svg/heatmap_subset.svg"))
heatmap.2(rhoMatrix, cellnote=sigMatrix,notecex=1.5,notecol="black",
          col=rev(redblue(40)), density.info="none", trace="none", 
          dendrogram=c("column"), symm=F,symkey=T,symbreaks=T, scale="none", 
          key.title = "", srtRow=45, adjRow=c(0, 1), srtCol=45, adjCol=c(1,1), 
          breaks=(-20:20)/20, margins = c(8, 8), cexRow=1.5, cexCol=1.5,font=2)

dev.off()
}
```
